---
title: "Final_project"
author: "Skasko_Stephen"
date: "11/26/2023"
output: html_document
---
Libraries 
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(glmnet)
library(BradleyTerry2)

```


```{r}
# Read in the files/data
nfl = read.csv("nfl_data.csv")
elo = read.csv("nfl_initial_elos.csv")

```

# Number of divisions 
```{r}
num_divisions <- length(unique(na.omit(elo$division)))
divisions <- unique(na.omit(elo$division))

print(num_divisions)
print(divisions)

```

# Relevent Statistics
```{r}
fgp <- read.csv("fgp.csv")
cp <- read.csv("cp.csv")

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Remove percentage sign and convert columns to numeric (handling NAs)
cp$Cmp. <- as.numeric(gsub("%", "", as.character(cp$Cmp.), fixed = TRUE), na.rm = TRUE)
fgp$FG. <- as.numeric(gsub("%", "", as.character(fgp$FG.), fixed = TRUE), na.rm = TRUE)

# Create a horizontal bar plot for completion percentage by team
ggplot(cp, aes(x = Cmp., y = reorder(Tm, Cmp.))) +
  geom_bar(stat = "identity", position = "dodge", color = "black", fill = "darkgreen") +
  labs(title = "Completion Percentage by Team",
       x = "Completion Percentage",
       y = "Team") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 0))

# Create a horizontal bar plot for field goal percentage by team
ggplot(fgp, aes(x = FG., y = reorder(Tm, FG.))) +
  geom_bar(stat = "identity", position = "dodge", color = "black", fill = "lightgreen") +
  labs(title = "Field Goal Percentage by Team",
       x = "Field Goal Percentage",
       y = "Team") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 0))


# Calculate mean and standard deviation for completion percentage
mean_completion <- mean(cp$Cmp., na.rm = TRUE)
sd_completion <- sd(cp$Cmp., na.rm = TRUE)

# Calculate mean and standard deviation for field goal percentage
mean_fg <- mean(fgp$FG., na.rm = TRUE)
sd_fg <- sd(fgp$FG., na.rm = TRUE)

# Display the results
cat("Completion Percentage:\n")
cat("Mean:", mean_completion, "\n")
cat("Standard Deviation:", sd_completion, "\n\n")

cat("Field Goal Percentage:\n")
cat("Mean:", mean_fg, "\n")
cat("Standard Deviation:", sd_fg, "\n")

```


# Simple Statistics - Team Performances, distribution, mean, standard deviation, extreme (high/low) values, outliers
```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Filter data for the year 2022
nfl_2022 <- nfl %>%
  filter(year == 2022)

# Calculate team statistics for the year 2022
team_stats_2022 <- nfl_2022 %>%
  group_by(home_team) %>%
  summarise(
    mean_home_score = mean(home_score),
    sd_home_score = sd(home_score),
    max_home_score = max(home_score),
    min_home_score = min(home_score)
  )

ggplot(team_stats_2022, aes(x = home_team, y = mean_home_score)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Mean Home Scores for Each Team in 2022",
       x = "Teams", y = "Mean Home Scores") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(team_stats_2022, aes(x = home_team, y = sd_home_score)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Standard Deviation of Home Scores for Each Team in 2022",
       x = "Teams", y = "Standard Deviation of Home Scores") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(team_stats_2022, aes(x = home_team, y = max_home_score)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(title = "Maximum Home Scores for Each Team in 2022",
       x = "Teams", y = "Maximum Home Scores") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(team_stats_2022, aes(x = home_team, y = min_home_score)) +
  geom_bar(stat = "identity", fill = "lightsalmon") +
  labs(title = "Minimum Home Scores for Each Team in 2022",
       x = "Teams", y = "Minimum Home Scores") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(team_stats_2022)



```

# Scoring Distribution FIXED
```{r}
# Scoring Distribution - Home Scores
home_scores <- nfl_scores$home_score
hist(home_scores, probability = TRUE, xlab = "Home Points Scored", main = "Home Scores Distribution")
curve(dnorm(x, mean = mean(home_scores), sd = sd(home_scores)), col = "blue", lwd = 2, add = TRUE)
qqnorm(home_scores, pch = 19, main = "Normal Probability Plot - Home Scores")
qqline(home_scores)
shapiro.test(home_scores)

# Scoring Distribution - Away Scores
away_scores <- nfl_scores$away_score
hist(away_scores, probability = TRUE, xlab = "Away Points Scored", main = "Away Scores Distribution")
curve(dnorm(x, mean = mean(away_scores), sd = sd(away_scores)), col = "green", lwd = 2, add = TRUE)
qqnorm(away_scores, pch = 19, main = "Normal Probability Plot - Away Scores")
qqline(away_scores)
shapiro.test(away_scores)

# Scatterplot - Home vs. Away Scores
plot(home_scores, away_scores, main = "Scatterplot - Home vs. Away Scores", xlab = "Home Scores", ylab = "Away Scores", pch = 19)

# Summary statistics of home and away scores
cat("Home Scores - Mean:", mean(home_scores), "SD:", sd(home_scores), "\n")
cat("Away Scores - Mean:", mean(away_scores), "SD:", sd(away_scores), "\n")
cat("Correlation between Home and Away Scores:", cor(home_scores, away_scores), "\n")


# Margin of Victory Calculation
nfl_scores$margin_of_victory <- nfl_scores$home_score - nfl_scores$away_score

# Plot - Margin of Victory
hist(nfl_scores$margin_of_victory, probability = TRUE, 
     xlab = "Margin of Victory (Home Score - Away Score)", 
     main = "Margin of Victory Distribution")
curve(dnorm(x, mean = mean(nfl_scores$margin_of_victory), sd = sd(nfl_scores$margin_of_victory)), 
      col = "orange", lwd = 2, add = TRUE)
qqnorm(nfl_scores$margin_of_victory, pch = 19, main = "Normal Probability Plot - Margin of Victory")
qqline(nfl_scores$margin_of_victory)
shapiro.test(nfl_scores$margin_of_victory)

# Summary statistics of margin of victory
cat("Margin of Victory - Mean:", mean(nfl_scores$margin_of_victory), "SD:", sd(nfl_scores$margin_of_victory), "\n")

```


# Calculating Pythagorean  - Win Percentage for each team, comparing win percentage vs actual win percentage plot, and correlation between the win percentage and actual win percentage
```{r}
scores <- read.table("nfl_data.csv", header = TRUE, sep = ",")
scores <- scores[scores$season == 2022 & scores$game_type == "r", ]


# Assuming 'scores' is your original data frame with game-level information
# Create a new data frame to store aggregated information for each team
team_data <- data.frame(
  team = character(),
  season = integer(),
  wins = integer(),
  losses = integer(),
  scored = integer(),
  allowed = integer()
)

# Iterate through each row in the original data frame
for (i in 1:nrow(scores)) {
  home_team <- scores$home_team[i]
  away_team <- scores$away_team[i]
  season <- scores$season[i]
  home_score <- scores$home_score[i]
  away_score <- scores$away_score[i]

  # Update home team's information
  home_team_row <- team_data[team_data$team == home_team & team_data$season == season, ]
  if (nrow(home_team_row) == 0) {
    home_team_row <- data.frame(
      team = home_team,
      season = season,
      wins = 0,
      losses = 0,
      scored = 0,
      allowed = 0
    )
    team_data <- rbind(team_data, home_team_row)
  }

  team_data$wins[team_data$team == home_team & team_data$season == season] <- team_data$wins[team_data$team == home_team & team_data$season == season] +
    ifelse(home_score > away_score, 1, 0)
  team_data$losses[team_data$team == home_team & team_data$season == season] <- team_data$losses[team_data$team == home_team & team_data$season == season] +
    ifelse(home_score < away_score, 1, 0)
  team_data$scored[team_data$team == home_team & team_data$season == season] <- team_data$scored[team_data$team == home_team & team_data$season == season] + home_score
  team_data$allowed[team_data$team == home_team & team_data$season == season] <- team_data$allowed[team_data$team == home_team & team_data$season == season] + away_score

  # Update away team's information
  away_team_row <- team_data[team_data$team == away_team & team_data$season == season, ]
  if (nrow(away_team_row) == 0) {
    away_team_row <- data.frame(
      team = away_team,
      season = season,
      wins = 0,
      losses = 0,
      scored = 0,
      allowed = 0
    )
    team_data <- rbind(team_data, away_team_row)
  }

  team_data$wins[team_data$team == away_team & team_data$season == season] <- team_data$wins[team_data$team == away_team & team_data$season == season] +
    ifelse(away_score > home_score, 1, 0)
  team_data$losses[team_data$team == away_team & team_data$season == season] <- team_data$losses[team_data$team == away_team & team_data$season == season] +
    ifelse(away_score < home_score, 1, 0)
  team_data$scored[team_data$team == away_team & team_data$season == season] <- team_data$scored[team_data$team == away_team & team_data$season == season] + away_score
  team_data$allowed[team_data$team == away_team & team_data$season == season] <- team_data$allowed[team_data$team == away_team & team_data$season == season] + home_score
}

# Print the resulting data frame
print(team_data)


# Calculate Pythagorean Win Percentage
team_data$pyth_wpct <- team_data$scored^2 / (team_data$scored^2 + team_data$allowed^2)

# Calculate actual Win Percentage
team_data$actual_wpct <- team_data$wins / (team_data$wins + team_data$losses)

# Plot comparison
plot(team_data$pyth_wpct, team_data$actual_wpct, pch = 19, xlab = "Pythagorean Win Percentage", ylab = "Actual Win Percentage", main = "Pythagorean vs Actual Win Percentage")

# Add team names to the plot with jitter
text(jitter(team_data$pyth_wpct), jitter(team_data$actual_wpct), labels = team_data$team, pos = 3, cex = 0.7)

# Calculate correlation
correlation <- cor(team_data$pyth_wpct, team_data$actual_wpct)
print(paste("Correlation:", correlation))

View(team_data)
```

# Bradley-Terry Model  - Show model if scoring distribution is normal, if low-scoring. Then ratings and coefficients for each team
```{r}
nfl_scores <- read.table("nfl_data.csv", header = TRUE, sep = ",")
nfl_scores <- nfl_scores[nfl_scores$season == 2022 & nfl_scores$game_type == "r", ]
nfl_scores$home_team <- gsub(" ", ".", nfl_scores$home_team)
nfl_scores$away_team <- gsub(" ", ".", nfl_scores$away_team)



nfl_data = as.data.frame(matrix(nrow = 271, ncol = 34))

teams = sort(unique(nfl_scores$home_team))

colnames(nfl_data) = c("MoV", "Location", teams)

nfl_data[is.na(nfl_data)] = 0

for (i in 1:length(teams)) {  
    home = as.integer(nfl_scores$home_team == teams[i]) 
    away = -1*as.integer(nfl_scores$away_team == teams[i]) 
    team_locations = home + away 
    nfl_data[, i+2] = team_locations 
}

nfl_data$MoV = nfl_scores$home_score - nfl_scores$away_score
nfl_data$Location = nfl_scores$location

colnames(nfl_data) <- make.unique(colnames(nfl_data))



# Fit Bradley-Terry model
model <- lm(MoV ~ 0 + ., data = nfl_data)

# Extract coefficients
coefficients <- coef(model, complete = TRUE)
coefficients[length(coefficients)] <- 0

# Adjust ratings and create a table
adjustment <- -sum(coefficients[2:length(coefficients)]) / (length(coefficients) - 1)
ratings <- c(coefficients[1], coefficients[2:length(coefficients)] + adjustment)

# Display ratings table
ratings_table <- data.frame(team = colnames(nfl_data)[-c(1, 2)], rating = ratings[-1])
print(ratings_table)

# Plot residuals or histogram (choose one, comment out the other)
residuals <- residuals(model)

# Choose either residual plot or histogram (comment out as needed)
# plot(predicted, residuals, pch = 19, xlab = "Predicted Margin of Victory", ylab = "Residual")
# abline(h = 0, col = "red")  # Add a horizontal line at y = 0

hist(residuals, xlab = "Residual", breaks = 12)
# qqnorm(residuals, pch = 19)
# qqline(residuals)

nfl_scores$home_win_prob <- 1 - pnorm((0 - ((coefficients[nfl_scores$home_team] + coefficients[1]) - coefficients[nfl_scores$away_team])) / sigma(model), 0, 1)

hist(nfl_scores$home_win_prob, xlab = "Home Win Probability", main = "Histogram of NFL Win Probability ")


# Identify overachieving, average, and underachieving teams
overachieving_teams <- ratings_table[ratings_table$rating > mean(ratings_table$rating), ]
average_teams <- ratings_table[ratings_table$rating <= mean(ratings_table$rating) & ratings_table$rating >= median(ratings_table$rating), ]
underachieving_teams <- ratings_table[ratings_table$rating < median(ratings_table$rating), ]

# Remove dots from team names
overachieving_teams$team <- gsub("\\.", " ", overachieving_teams$team)
average_teams$team <- gsub("\\.", " ", average_teams$team)
underachieving_teams$team <- gsub("\\.", " ", underachieving_teams$team)

library(kableExtra)

kable(overachieving_teams, caption = "Overachieving Teams", row.names = FALSE) %>%
  kable_styling()

kable(average_teams, caption = "Average Teams", row.names = FALSE) %>%
  kable_styling()

kable(underachieving_teams, caption = "Underachieving Teams", row.names = FALSE) %>%
  kable_styling()


```


# Preseaosn and Postseason Elo Ratiing - elo rating for each team before and after the season,summary + comparison of preseason vs postseason
```{r}
elos = read.table("nfl_initial_elos.csv", header = TRUE, sep = ",")
scores = read.table("nfl_data.csv", header = TRUE, sep = ",")

home_wins = 0
games = 0
first_game_index = 271


# Iterate through games - first index can be changed to eliminate early seasons where scores are extreme
for(i in first_game_index:nrow(scores)) {  
  # Count number of games that do not end in ties
  if(scores$home_score[i] != scores$away_score[i] && scores$location[i] == 1) { games = games + 1 }
  
  # Count number of games where home team wins
  if(scores$home_score[i] > scores$away_score[i] && scores$location[i] == 1) { home_wins = home_wins + 1 }
}

home_win_prob = home_wins / games  # Calculate home win probability where outcome was not a tie
hfa = -400*log10(1/home_win_prob - 1)  # Calculate number of Elo points added to home team
elos = read.table("nfl_initial_elos.csv", header = TRUE, sep = ",")
scores = read.table("nfl_data.csv", header = TRUE, sep = ",")

starting_weight = 0  # Lower bound for weight ranges to be tested - generally set equal to 0
iterations = 100  # Number of k values to test
step_size = 0.1  # Amount to increment k by at each step
first_game_index = 271

if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}


# Initialize data frame to store k values and corresponding error
errors = data.frame(matrix(ncol = 2, nrow = iterations))
colnames(errors) = c("weight", "error")
errors$weight = starting_weight + (1:iterations)*step_size
errors$error = NA

# Iterate through all potential k values that are being tested
for(counter in 1:iterations) {
    weight = starting_weight + counter*step_size  # Calculate k value for current iteration
    print(weight)
    error = 0  # Reset error for current iteration
    elos = read.table("nfl_initial_elos.csv", header = TRUE, sep = ",")  # Reset initial Elo ratings

    # Iterate through games - first index can be changed to eliminate early seasons in a league where early results tend to be extreme
    for(i in first_game_index:nrow(scores)) {  
        # Find indices corresponding to home and away teams for current game
        home_index = which(elos$team == scores$home_team[i])
        away_index = which(elos$team == scores$away_team[i])
        
        # Find home and away team Elo ratings
        home_elo = elos$rating[home_index]
        away_elo = elos$rating[away_index]
        
        # Calculate home team win probability
        win_prob = 1 / (10^((away_elo - (home_elo + hfa*scores$location[i]))/400) + 1)
        
        # Calculate actual margin of victory - must be positive
        score_diff = abs(scores$home_score[i] - scores$away_score[i])  
        
        # Determine home team result
        if(scores$home_score[i] > scores$away_score[i]) { 
          home_result = 1  # Home team wins
        } else if(scores$home_score[i] < scores$away_score[i]) { 
          home_result = 0  # Home team loses
        } else { 
          home_result = 0.5  # Tie
        }
        
        # Add squared error between home result and predicted probability of home team winning to SSE
        error = error + (home_result - win_prob)^2
        
        # Calculate amount each team's Elo rating is adjusted by
        home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)   # Log scaling function

        # Adjust Elo ratings - add point to winner and subtract points from loser
        elos$rating[home_index] = elos$rating[home_index] + home_elo_adjustment
        elos$rating[away_index] = elos$rating[away_index] - home_elo_adjustment

    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
        if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {
            for(j in 1:nrow(elos)) {
                if(scores$season[i] >= elos$inaugural_season[j]) {
                    elos$rating[j] = elos$rating[j] - (elos$rating[j] - 1500)/3
                }
            }
    
            existing_teams = elos[which(elos$inaugural_season <= (scores$season[i] + 1)),]
            expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
    
            for(j in 1:nrow(elos)) {
                if((scores$season[i] + 1) >= elos$inaugural_season[j]) {
                    elos$rating[j] = elos$rating[j] + expansion_adjustment
                }
            }
        }
    }
    errors$error[counter] = error  # Store error for current iteration
}

# Create plot of SSE against k-values
plot(errors$weight, errors$error, xlab = "k Value", ylab = "Sum of Squared Errors")

# Choose and print optimal weight based on value that had the lowest SSE
weight = errors$weight[which(errors$error == min(errors$error))]
print(weight)

plot(errors$weight, errors$error, pch = 19, xlab = "Weight", ylab = "Error")


# Remove rows with missing values
elo_comparison <- elo_comparison[complete.cases(elo_comparison[, c("Beginning_Elo", "End_Elo", "Team")]), ]

# Scatterplot with text labels and pointers using ggplot
ggplot(elo_comparison, aes(x = Beginning_Elo, y = End_Elo, label = Team)) +
  geom_point(size = 3) +
  geom_text(aes(label = Team), hjust = -0.1, vjust = 1) +
  labs(title = "Beginning vs. End of Season Elo Ratings",
       x = "Beginning of Season Elo Rating",
       y = "End of Season Elo Rating") +
  theme_minimal() +
  xlim(1400, 1650)  # Set x-axis limits


# Load knitr package
library(knitr)

# Elo Ratings Table
elo_table <- results[, c("team", "starting_elo", "ending_elo")]

# Print the table using kable
kable(elo_table, format = "markdown", col.names = c("Team", "Starting Elo", "Ending Elo"))




```

```{r}
elo_table <- results[, c("team", "starting_elo", "ending_elo")]

# Print the table using kable
kable(elo_table, format = "markdown", col.names = c("Team", "Starting Elo", "Ending Elo"))

# Assuming elos_diff_2022 is your data frame with Elo rating differences for 32 teams
# Adjust the threshold as needed for playoff qualification

# Identify the best and worst teams
best_team <- results[which.max(results$Elo_Diff), ]
worst_team <- results[which.min(results$Elo_Diff), ]

# Print best and worst teams
cat("Best Team:\n")
print(best_team)
cat("\nWorst Team:\n")
print(worst_team)



```


# One Team - selecting one team and create a graph/table of their elo ratings over time
```{r}
# Select team and season to follow over a specified period of time
team = "Pittsburgh Steelers"
first_season = 2022
last_season = 2022


# Input the optimal k factors (weight) and home field advantage
weight = 10
hfa = 53.67


# Identify if ties exist in the sport, which determines if an additional 0.5 needs to be added to the margin of victory
if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}

# Create data frame to store information for team specified above
team_results = data.frame(matrix(ncol = 10, nrow = 0))
colnames(team_results) = c("opponent", "pregame_elo", "opponent_pregame_elo", "win_probability", "result", "team_score", "opponent_score", "elo_adjustment", "postgame_elo", "opponent_postgame_elo")

# Iterate through all games in the sport's history
for(i in 1:nrow(scores)) {
    # Find indices corresponding to home and away teams for current game
    home_index = which(elos$team == scores$home_team[i])
    away_index = which(elos$team == scores$away_team[i])
    
    
    # Find home and away team Elo ratings
    home_elo = elos$rating[home_index]
    away_elo = elos$rating[away_index]
    
    
    # Calculate home team win probability
    win_prob = 1 / (10^((away_elo - (home_elo + hfa*scores$location[i]))/400) + 1)
    
    
    # Calculate actual margin of victory - must be positive
    score_diff = abs(scores$home_score[i] - scores$away_score[i])   
    
    
    # Determine home team result
    if(scores$home_score[i] > scores$away_score[i]) { 
        home_result = 1  # Home team wins
    } else if(scores$home_score[i] < scores$away_score[i]) { 
        home_result = 0  # Home team loses
    } else { 
        home_result = 0.5  # Tie
    }
  
    
    # Calculate amount each team's Elo rating is adjusted by
    home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)
    
    
    # Adjust Elo ratings - add point to winner and subtract points from loser
    elos$rating[home_index] = elos$rating[home_index] + home_elo_adjustment
    elos$rating[away_index] = elos$rating[away_index] - home_elo_adjustment

    
    # Add game information to team result data frame for each team game of the team specified above if team and season both match
    if(scores$season[i] >= first_season & scores$season[i] <= last_season & (scores$home_team[i] == team | scores$away_team[i] == team)) {
        if(scores$home_team[i] == team) {  # If specified team was at home
            team_results[nrow(team_results) + 1,] = c(scores$away_team[i], elos$rating[home_index] - home_elo_adjustment, elos$rating[away_index] + home_elo_adjustment, win_prob, home_result, scores$home_score[i], scores$away_score[i], home_elo_adjustment, elos$rating[home_index], elos$rating[away_index])
        } else {  # If specified team was away
            team_results[nrow(team_results) + 1,] = c(scores$home_team[i], elos$rating[away_index] + home_elo_adjustment, elos$rating[home_index] - home_elo_adjustment, 1-win_prob, 1-home_result, scores$away_score[i], scores$home_score[i], -1*home_elo_adjustment, elos$rating[away_index], elos$rating[home_index])
        }
    }
    
    
    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
    if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {  # New season
        for(j in 1:nrow(elos)) {  # For each team
            if(scores$season[i] >= elos$inaugural_season[j]) {  # Check if team existed
                # Move each team's Elo rating back towards 1500 by 1/3 of the difference
                elos$rating[j] = elos$rating[j] - (elos$rating[j] - 1500)/3
            }
        }
        
        
        # Identify all teams that existed at beginning of following season
        existing_teams = elos[which(elos$inaugural_season <= (scores$season[i] + 1)),]
        
        
        # Calculate amount each team's Elo rating must be adjusted by to make mean 1500
        expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
        
        
        # Perform expansion adjustment on teams that existed at beginning of following season
        for(j in 1:nrow(elos)) {  # For each team
            if((scores$season[i] + 1) >= elos$inaugural_season[j]) {  # Check if team existed
                elos$rating[j] = elos$rating[j] + expansion_adjustment  # Update ratings if so
            }
        }
    }
}


# Create plot of postgame Elo ratings
if (first_season == last_season) {
    team_name = paste(first_season, team)
} else {
    team_name = paste(paste(first_season, last_season, sep = "-"), team)
}

plot(team_results$postgame_elo, type = "l", xlab = team_name, ylab = "Elo Rating")
abline(h = 1500)
```

# Monte Carlo - monte carlo simulation for elo ratings, actual/simulated win totals and division titles, then create histogram/scatterplot of residuals comparing  actual and simulated win totals
```{r}
# Specify season to simulate
simulated_season = 2022

# Read in list of all games
scores = read.table("nfl_data.csv", header = TRUE, sep = ",")

# Read in initial Elo rating, conference, and division
team_info = read.table("nfl_initial_elos.csv", header = TRUE, sep = ",")

# Obtain list of unique conference names and unique division names
conferences = na.omit(unique(team_info$conference))
divisions = na.omit(unique(team_info$division))

# Create list of games that occurred prior to season being simulated
pre_season = scores[which(scores$season < simulated_season & scores$season >= 1901),]

# Create list of regular season games for season being simulated
season_schedule = scores[which(scores$season == simulated_season & (scores$game_type == "r" | scores$game_type == "np")),]

# Input the optimal k factors (weight) and home field advantage
hfa = 53.67
weight = 10

# Identify if ties exist in the sport, which determines if an additional 0.5 needs to be added to the margin of victory
if (sum(scores$tie) > 0) {
    tie_adj = 0.5 
} else {
    tie_adj = 0
}

# Iterate through all games in the sport's history up to season being simulated
for(i in 1:nrow(pre_season)) {
    # Find indices corresponding to home and away teams for current game
    home_index = which(team_info$team == pre_season$home_team[i])
    away_index = which(team_info$team == pre_season$away_team[i])
    
    # Find home and away team Elo ratings
    home_elo = team_info$rating[home_index]
    away_elo = team_info$rating[away_index]
    
    # Calculate home team win probability
    win_prob = 1 / (10^((away_elo - (home_elo + hfa*pre_season$location[i]))/400) + 1)
    
    # Calculate actual margin of victory - must be positive
    score_diff = abs(pre_season$home_score[i] - pre_season$away_score[i])
    
    # Determine home team result
    if(pre_season$home_score[i] > pre_season$away_score[i]) { 
    home_result = 1  # Home team wins
    } else if(pre_season$home_score[i] < pre_season$away_score[i]) { 
    home_result = 0  # Home team loses
    } else { 
    home_result = 0.5  # Tie
    }
    
    # Calculate amount each team's Elo rating is adjusted by
    home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)

  
    # Adjust Elo ratings - add point to winner and subtract points from loser
    team_info$rating[home_index] = team_info$rating[home_index] + home_elo_adjustment
    team_info$rating[away_index] = team_info$rating[away_index] - home_elo_adjustment
    
    # Adjust Elo ratings at end of season to regress 1/3 of the way towards 1500
    if(i < nrow(scores) && scores$season[i+1] > scores$season[i]) {
        for(j in 1:nrow(team_info)) {
            if(scores$season[i] >= team_info$inaugural_season[j]) {
                team_info$rating[j] = team_info$rating[j] - (team_info$rating[j] - 1500)/3
                }
        }
    
        # Identify all teams that existed at beginning of following season
        existing_teams = team_info[which(team_info$inaugural_season <= (scores$season[i] + 1)),]
    
        # Calculate amount each team's Elo rating must be adjusted by to make mean 1500
        expansion_adjustment = -1*(mean(existing_teams$rating) - 1500)
    
        # Perform expansion adjustment on teams that existed at beginning of following season
        for(j in 1:nrow(team_info)) {
            if((scores$season[i] + 1) >= team_info$inaugural_season[j]) {
                team_info$rating[j] = team_info$rating[j] + expansion_adjustment
            }
        }
    }
}

# Set seed for replication purposes
set.seed(31)

# Determine number of times to simulate the season
iterations = 10000

# Create data frame to hold Elo ratings, actual win totals, and simulation results
results = data.frame(matrix(0, ncol = 6, nrow = nrow(team_info)))
colnames(results) = c("team", "starting_elo", "ending_elo", "actual_wins", "average_wins", "division_titles")
results$team = team_info$team

# Create data frame to hold number of wins by each team in each iteration
win_totals = data.frame(matrix(0, ncol = nrow(team_info), nrow = iterations))
colnames(win_totals) = team_info$team

# Simulate the season the given number of times
for(i in 1:iterations) {
    if(i %% 1000 == 0) {print(i)}
    season_stats = team_info[,which(colnames(team_info) != "inaugural_season")]
    season_stats$wins = 0
    season_stats$rand = runif(nrow(team_info))
    
    # Simulate each game in current season
    for(j in 1:nrow(season_schedule)) {
        # Find indices corresponding to home and away teams for current game
        home_index = which(season_stats$team == season_schedule$home_team[j])
        away_index = which(season_stats$team == season_schedule$away_team[j])
        
        # Find home and away team Elo ratings
        home_elo = season_stats$rating[home_index]
        away_elo = season_stats$rating[away_index]
        
        # Calculate home team win probability
        win_prob = 1 / (10^((away_elo - (home_elo + hfa*season_schedule$location[j]))/400) + 1)
        u = runif(1)  # Generate a random number used to determine the winner of the game
        
        # Determine which team wins the simulated game and increment their win total by 1
        if(u < win_prob) {
            season_stats$wins[home_index] = season_stats$wins[home_index] + 1
        } else {
            season_stats$wins[away_index] = season_stats$wins[away_index] + 1
        }
    
        # Calculate actual margin of victory - must be positive
        score_diff = abs(season_schedule$home_score[j] - season_schedule$away_score[j])
        
        # Determine home team result
        if(season_schedule$home_score[j] > season_schedule$away_score[j]) { 
          home_result = 1  # Home team wins
        } else if(season_schedule$home_score[j] < season_schedule$away_score[j]) { 
          home_result = 0  # Home team loses
        } else { 
          home_result = 0.5  # Tie
        }
        
        # Calculate amount each team's Elo rating is adjusted by
        home_elo_adjustment = weight * log(score_diff + 1 + tie_adj) * (home_result - win_prob)
        
        # Adjust Elo ratings after game has been simulated to get team's new strength
        season_stats$rating[home_index] = season_stats$rating[home_index] + home_elo_adjustment
        season_stats$rating[away_index] = season_stats$rating[away_index] - home_elo_adjustment
    
        # Track season wins
        if(i == 1) {
            if(season_schedule$home_score[j] > season_schedule$away_score[j]) {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 1
            } else if(season_schedule$home_score[j] < season_schedule$away_score[j]) {
                results$actual_wins[away_index] = results$actual_wins[away_index] + 1
            } else {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 0.5
                results$actual_wins[away_index] = results$actual_wins[away_index] + 0.5
            }
        }
    }

    # Define data frame that contains division winners
    division_winners = data.frame(matrix(ncol = 6, nrow = 0))
    colnames(division_winners) = c("team", "conference", "division", "rating", "wins", "rand")
  
    # For each division
    for(div in divisions) {
        div_standings = season_stats[which(season_stats$division == div),]  # Identify all teams in current division
        div_standings = div_standings[order(-div_standings$wins, -div_standings$rand),]  # Sort division by wins and random number
        division_winners = rbind(division_winners, div_standings[1,])  # Add division winner to 'division_winners' data frame
    }
  
    # Save end of season win totals (from column 5 of season_stats) for each iteration in the win_totals data frame
    win_totals[i,] = t(season_stats[5])
    
    # Add number of wins for each team during this iteration to sum
    results$average_wins = results$average_wins + season_stats$wins
  
    # Increment the number of division titles for each division winner by 1
    for(team in division_winners$team) {
        index = which(season_stats$team == team)  # Index of division winner
        results$division_titles[index] = results$division_titles[index] + 1  # Increment division titles
    }
}
 
# Calculate average number of wins across all iterations
results$starting_elo = team_info$rating
results$ending_elo = season_stats$rating
results$average_wins = results$average_wins/iterations
results$division_titles = results$division_titles/iterations

# Plots of preseason and postseason Elo rating against simulated wins
plot(results$starting_elo[1:32], results$average_wins[1:32], xlab = "Preseason Elo Rating", ylab = "Average Wins", pch = 19)
plot(results$ending_elo[1:32], results$average_wins[1:32], xlab = "Postseason Elo Rating", ylab = "Average Wins", pch = 19)

# Plot of change in Elo rating against win residual
plot(results$ending_elo[1:32] - results$starting_elo[1:32], results$actual_wins[1:32] - results$average_wins[1:32], xlab = "Postseason Elo Rating", ylab = "Actual Wins - Average Wins", pch = 19)
abline(h = 0)
abline(v = 0)

# Plots of preseason and postseason Elo rating against simulated division titles
plot(results$starting_elo, results$division_titles, xlab = "Preseason Elo Rating", ylab = "Proportion of Division Titles", pch = 19)
abline(v = 1540)
abline(h = .40)

plot(results$ending_elo, results$division_titles, xlab = "Postseason Elo Rating", ylab = "Proportion of Division Titles", pch = 19)
abline(v = 1525)
abline(h = .40)




# Assuming 'team_data' is your data frame with actual and simulated win totals
residuals <- team_data$actual_wpct - team_data$pyth_wpct

# Create histogram of residuals
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals", col = "skyblue")

# Assuming 'team_data' is your data frame with actual and simulated win totals
plot(team_data$actual_wpct, team_data$pyth_wpct, pch = 19, col = "black",
     xlab = "Actual Win Percentage", ylab = "Simulated Win Percentage",
     main = "Scatterplot of Actual vs. Simulated Win Percentages")

# Add team names to the plot
text(team_data$actual_wpct, team_data$pyth_wpct, labels = team_data$team, pos = 3, cex = 0.8, col = "black")
```

```{r}
# Combine actual and simulated win totals into a single data frame
team_data <- data.frame(
  team = team_info$team,
  actual_wins = results$actual_wins,
  simulated_wins = results$average_wins,
  division_titles = results$division_titles
)

# Table of actual and simulated win totals, and number of division titles
team_table <- team_data[, c("team", "actual_wins", "simulated_wins", "division_titles")]
print(knitr::kable(team_table, format = "markdown", col.names = c("Team", "Actual Wins", "Simulated Wins", "Division Titles")))




# Combine actual and simulated win totals into a single data frame
team_data <- data.frame(
  team = team_info$team,
  actual_wins = results$actual_wins,
  simulated_wins = results$average_wins,
  division_titles = results$division_titles
)

# Print the table in a format suitable for viewing in the R console
print(team_data[, c("team", "actual_wins", "simulated_wins", "division_titles")])


```

# Overahcieving and Underachieving teams  histogram of their simulated win/point totals
```{r}

selected_team <- "Buffalo Bills"  

team_wins <- win_totals[, selected_team]

hist(team_wins, 
     main = paste("Histogram of Win Totals for", selected_team),
     xlab = "Win Totals",
     ylab = "Frequency",
     col = "darkgreen",
     border = "black")
legend("topright", legend = selected_team, col = "darkgreen", pch = 15)

selected_team <- "Houston Texans"  

team_wins <- win_totals[, selected_team]

hist(team_wins, 
     main = paste("Histogram of Win Totals for", selected_team),
     xlab = "Win Totals",
     ylab = "Frequency",
     col = "darkgreen",
     border = "black")
legend("topright", legend = selected_team, col = "darkgreen", pch = 15)



# Assuming you have additional data frames for points, quarterback performance, etc.

# Add necessary libraries if not already loaded
library(dplyr)

# Function to calculate additional metrics
calculate_additional_metrics <- function(scores, team_name) {
  team_data <- filter(scores, home_team == team_name | away_team == team_name)
  
  # Calculate point differentials
  team_data <- mutate(team_data, point_diff = ifelse(home_team == team_name,
                                                      home_score - away_score,
                                                      away_score - home_score))
  
  # Add more calculations as needed
  
  return(team_data)
}

# Apply the function to the selected teams
buffalo_data <- calculate_additional_metrics(scores, "Buffalo Bills")
houston_data <- calculate_additional_metrics(scores, "Houston Texans")

# Plot histograms of additional metrics (e.g., point differentials)
hist(buffalo_data$point_diff, 
     main = paste("Histogram of Point Differentials for Buffalo Bills"),
     xlab = "Point Differentials",
     ylab = "Frequency",
     col = "blue",
     border = "black")
legend("topright", legend = "Buffalo Bills", col = "blue", pch = 15)

hist(houston_data$point_diff, 
     main = paste("Histogram of Point Differentials for Houston Texans"),
     xlab = "Point Differentials",
     ylab = "Frequency",
     col = "red",
     border = "black")
legend("topright", legend = "Houston Texans", col = "red", pch = 15)

```

